---
title: "growth_analysis"
author: "Denver Link"
date: "2025-06-23"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Packages
```{r}
library(tidyverse)
library(arrow)
library(ggplot2)
library(dplyr)
library(sf)
library(ggspatial)
library(ggthemes)
library(rnaturalearth)
library(rnaturalearthdata)
library(leaflet)
```

# Data
-growth data from MN
-zm and ssw lakes from mn infested waters list
```{r}
#reading in the updated Minnesota data
mn_data <- open_dataset(sources = file.path("D:",
                                            "Shared drives",
                                            "Hansen Lab",
                                            "RESEARCH PROJECTS",
                                            "Fish Survey Data",
                                            "Parquet files",
                                            "mn_update")) %>% 
  filter(state == "Minnesota" & !is.na(est_age))

zm_ssw <- read_csv(file.path("D:",
                             "Shared drives",
                             "Hansen Lab",
                             "RESEARCH PROJECTS",
                             "MAISRC Mercury Pathways 53.2",
                             "Data",
                             "zm_ssw_6jan2026.csv")) %>% 
  #filter out na dows or those "none"
  filter(!is.na(dow) & !(str_detect(dow, "n"))) %>% 
  #generate original id for reference
  mutate(original_id = dow) %>% 
  #remove dashes
  #make sure short dows are 8 digits
  #fix a few manually
  #remove leading 0s to match the fish data
  mutate(dow = str_replace_all(dow, pattern = "-", ""),
         dow = if_else(nchar(dow) == 6, paste0(dow, "00"), dow),
         dow = case_when(dow == "10179" ~ "01017900",
                         str_detect(dow, "040063") ~ "04006300",
                         TRUE ~ dow),
         dow = case_when(str_starts(dow, "0") ~ str_sub(dow, 2),
                         TRUE ~ dow))  %>% 
  #rename dow to match fishereis column
  rename(lake_id = dow) %>% 
  #pivot to lake per row, with zm and starry columns 
  pivot_wider(id_cols = c(lake_id, lake_name),
              names_from = ais,
              values_from = year_infested,
              values_fn = min) %>% 
  rename(zm = `zebra mussel`,
         ssw = `starry stonewort`)

#crosswalk - get lat and lon from fish data
xwalk <- open_dataset(sources = file.path("D:",
                                            "Shared drives",
                                            "Hansen Lab",
                                            "RESEARCH PROJECTS",
                                            "Fish Survey Data",
                                            "Parquet files",
                                            "mn_update")) %>% 
  select(state, latitude_lake_centroid, lake_id, longitude_lake_centroid) %>% 
  filter(state == "Minnesota") %>% 
  filter(!is.na(latitude_lake_centroid) & !is.na(lake_id)) %>% 
  distinct(lake_id, latitude_lake_centroid, longitude_lake_centroid) %>% 
  collect() 
```

# Exploring growth data
```{r}
#how many lakes do we have for each species at the year level?
mn_data %>% 
  filter(alk == "year") %>% 
  group_by(species_1) %>% 
  summarise(n_lakes = n_distinct(lake_id)) %>% 
  collect()

#how many lakes with lake-year data do we have for each invasive?
mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  group_by(lake_id) %>% 
  mutate(ais = case_when(!is.na(zm) & is.na(ssw) ~ "zm",
                         !is.na(ssw) & is.na(zm) ~ "ssw",
                         !is.na(ssw) & !is.na(zm) ~ "both",
                         is.na(ssw) & is.na(zm) ~ "none")) %>% 
  ungroup() %>% 
  distinct(lake_id, ais) %>% 
  group_by(ais) %>% 
  count() %>% 
  collect()

#what is the general sense for species?
check <- mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  group_by(lake_id, species_1) %>% 
  mutate(ais = case_when(!is.na(zm) & is.na(ssw) ~ "zm",
                         !is.na(ssw) & is.na(zm) ~ "ssw",
                         !is.na(ssw) & !is.na(zm) ~ "both",
                         is.na(ssw) & is.na(zm) ~ "none"),
         impact_zm = case_when(year >= zm ~ "after",
                               year < zm ~ "before",
                               TRUE ~ "NA"),
         impact_ssw = case_when(year >= ssw ~ "after",
                                year < ssw ~ "before",
                                TRUE ~ "NA")) %>% 
  ungroup() %>% 
  distinct(lake_id, ais, species_1, impact_zm, impact_ssw) %>% 
  collect()
#we can see here if we have before and after for each invasive - NA for where it is uninvaded for that species
glimpse(check)

#general on fish we have growth data
invasion_summary <- check %>%
  group_by(lake_id, species_1) %>%
  summarise(
    zm_before = any(impact_zm == "before"),
    zm_after  = any(impact_zm == "after"),
    ssw_before = any(impact_ssw == "before"),
    ssw_after  = any(impact_ssw == "after"),
    .groups = "drop"
  ) %>%
  mutate(
    zm_status = case_when(
      zm_before & zm_after ~ "both",
      zm_before & !zm_after ~ "before only",
      !zm_before & zm_after ~ "after only",
      !zm_before & !zm_after ~ "neither"
    ),
    ssw_status = case_when(
      ssw_before & ssw_after ~ "both",
      ssw_before & !ssw_after ~ "before only",
      !ssw_before & ssw_after ~ "after only",
      !ssw_before & !ssw_after ~ "neither"
    )
  )

invasion_counts <- invasion_summary %>%
  group_by(species_1) %>% 
  count(zm_status, ssw_status) %>% 
  pivot_longer(cols = c(zm_status, ssw_status),names_to = "ais",values_to = "timing") %>% 
  group_by(species_1, ais, timing) %>% 
  summarise(n = sum(n))

invasion_counts %>% 
  filter(timing != "neither") %>% 
  ggplot() +
  geom_col(aes(timing, n)) +
  facet_grid(species_1~ais)

invasion_counts %>% 
  filter(timing != "neither") %>% 
  #filter(ais == "zm_status") %>% 
  ggplot() +
  geom_col(aes(timing, n)) +
  facet_grid(ais~species_1)

#how many sssw lakes do we have pre and post?
invasion_counts %>% 
  filter(ais == "ssw_status" & timing == "both")

ssw_lakes <- check %>% 
  filter(ais %in% c("ssw", "both")) %>% 
  select(-impact_zm, -ais) %>% 
  distinct(lake_id, species_1, impact_ssw) %>% 
  mutate(indicator = 1) %>% 
  pivot_wider(names_from = impact_ssw,
              values_from = indicator)
glimpse(ssw_lakes)

ssw_lakes_both <- ssw_lakes %>% 
  filter(before == "1" & after == "1") %>% 
  select(-after, -before) %>% 
  group_by(lake_id) %>% 
  summarise(species_1 = list(species_1), .groups = "drop") %>% 
  mutate(species_1 = sapply(species_1, toString))
  

###years of data?
years <- mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  group_by(lake_id, species_1, year) %>% 
  mutate(ais = case_when(!is.na(zm) & is.na(ssw) ~ "Zebra Mussel",
                         !is.na(ssw) & is.na(zm) ~ "Starry Stonewort",
                         !is.na(ssw) & !is.na(zm) ~ "ZM and SSW",
                         is.na(ssw) & is.na(zm) ~ "Uninvaded")) %>% 
  distinct(lake_id, ais, species_1, year) %>% 
  collect()

years %>% 
  group_by(species_1, lake_id, ais) %>% 
  count() %>% 
  ggplot() +
  geom_histogram(aes(n)) +
  labs(x = "# of years with halk at lake-year level",
       y = "# of lakes") +
  facet_grid(species_1~ais, scales = "free")
```

# Map (filtering for zebra mussel lakes with before and after data)
```{r}
lake_species <- mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  group_by(lake_id, species_1) %>% 
  mutate(ais = case_when(!is.na(zm) & is.na(ssw) ~ "zm",
                         !is.na(ssw) & is.na(zm) ~ "ssw",
                         !is.na(ssw) & !is.na(zm) ~ "both",
                         is.na(ssw) & is.na(zm) ~ "uninvaded"),
         impact_zm = case_when(year >= zm ~ "after",
                               year < zm ~ "before",
                               TRUE ~ "NA")) %>% 
  ungroup() %>% 
  distinct(lake_id, ais, species_1, impact_zm) %>% 
  collect() %>% 
  pivot_wider(id_cols = c(lake_id, species_1, ais), values_from = impact_zm,names_from = impact_zm) %>% 
  #zm_ssw has two a record for both zm and ssw for these lakes - leading an erroneous uninvaded row
  filter(!(lake_id == "18031100" & ais == "uninvaded") &
           !(lake_id == "4007900" & ais == "uninvaded")) %>% 
  #filter for cases where uninvaded lakes have data and invaded systems have pre and post
  mutate(timing = case_when(ais == "uninvaded" ~ "y",
                            ais == "ssw" ~ "n",
                            ais == "both" ~ "n",
                            ais == "zm" & !is.na(after) & !is.na(before) ~ "y",
                            TRUE ~ "n")) %>% 
  filter(timing == "y")

#we now attach spatial data, remove where they are na because they are mostly river or small
lake_species <- lake_species %>% 
  left_join(xwalk) %>% 
  filter(!is.na(latitude_lake_centroid)) %>% 
  select(lake_id, species_1, ais, latitude_lake_centroid, longitude_lake_centroid) %>% 
  #clean up species names 
  mutate(species_clean = case_when(species_1 == "black_crappie" ~ "Black Crappie",
                               species_1 == "bluegill" ~ "Bluegill",
                               species_1 == "cisco" ~ "Cisco",
                               species_1 == "largemouth_bass" ~ "Largemouth Bass",
                               species_1 == "northern_pike" ~ "Northern Pike",
                               species_1 == "smallmouth_bass" ~ "Smallmouth Bass",
                               species_1 == "walleye" ~ "Walleye",
                               species_1 == "yellow_perch" ~ "Yellow Perch"))

lake_sf <- lake_species %>%
  distinct(lake_id, species_clean, ais, latitude_lake_centroid, longitude_lake_centroid) %>%
  st_as_sf(coords = c("longitude_lake_centroid", "latitude_lake_centroid"), crs = 4326)

us_states <- ne_states(country = "United States of America", returnclass = "sf")
mn_border <- us_states %>% filter(name == "Minnesota")

# Plot
lake_uninvaded <- lake_sf %>% filter(ais == "uninvaded")
lake_zm <- lake_sf %>% filter(ais == "zm")

ggplot() +
  geom_sf(data = mn_border, fill = NA, color = "black") +
  geom_sf(data = lake_uninvaded, color = "lightblue", alpha = 0.5, size = 1) +
  geom_sf(data = lake_zm, color = "salmon", alpha = 0.75, size = 1) +
  facet_wrap(~ species_clean) +
  theme_minimal() +
  labs(color = "Invasion") +
  coord_sf(xlim = st_bbox(mn_border)[c("xmin", "xmax")],
           ylim = st_bbox(mn_border)[c("ymin", "ymax")]) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
#ggsave("all_spec_map.jpg", height = 5, width = 5)

#subset
lake_species_sub <- lake_species %>% 
  left_join(xwalk) %>% 
  filter(!is.na(latitude_lake_centroid)) %>% 
  select(lake_id, species_clean, species_1, ais, latitude_lake_centroid, longitude_lake_centroid) %>% 
  filter(species_clean %in% c("Walleye", "Northern Pike", "Bluegill"))

lake_sf <- lake_species_sub %>%
  distinct(lake_id, species_clean, species_1, ais, latitude_lake_centroid, longitude_lake_centroid) %>%
  st_as_sf(coords = c("longitude_lake_centroid", "latitude_lake_centroid"), crs = 4326)

lake_uninvaded <- lake_sf %>% filter(ais == "uninvaded")
lake_zm <- lake_sf %>% filter(ais == "zm")

ggplot() +
  geom_sf(data = mn_border, fill = NA, color = "black") +
  geom_sf(data = lake_uninvaded, color = "lightblue", alpha = 0.5, size = 1) +
  geom_sf(data = lake_zm, color = "salmon", alpha = 0.75, size = 1) +
  facet_wrap(~ species_clean) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.text = element_text(size = 12),
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(color = "Invasion") +
  coord_sf(xlim = st_bbox(mn_border)[c("xmin", "xmax")],
           ylim = st_bbox(mn_border)[c("ymin", "ymax")])
#ggsave("filtered_spec_map.jpg", height = 3, width = 5)

rm(check,
   invasion_counts,
   invasion_summary,
   lake_sf,
   lake_species,
   lake_uninvaded,
   lake_zm,
   mn_border,
   ssw_lakes,
   ssw_lakes_both,
   us_states,
   years)
```

# ZM, WAE-NOP-BLG raw fish data for growth
```{r}
age_data <- mn_data %>% 
  #filter MN data for fish with alked year level data
  filter(alk == "year") %>% 
  #filtering join to the lake-species combos we want
  right_join(lake_species_sub, by = c("lake_id", "species_1"), suffix = c("_fish", "_invasion")) %>% 
  collect() %>% 
  #selecting columns important from the fish data that are relevant 
  select(lake_id,
         nhdhr_id,
         date_total_effort_ident,
         date_sample,
         year,
         sampling_method,
         site_id,
         species_1,
         length_1,
         length_unit_1,
         age,
         aging_structure_1,
         est_age,
         alk,
         alk_age_str,
         alk_length,
         alk_n,
         weight_1,
         weight_unit_1,
         flag,
         notes_1) %>% 
  #bring in invasion data 
  left_join(zm_ssw, by = "lake_id") %>% 
  #create columns for invasion impact for both invaders
  mutate(impact_zm = case_when(year >= zm ~ "after",
                               year < zm ~ "before",
                               TRUE ~ NA),
         impact_zm_years = year - zm,
         impact_ssw = case_when(year >= ssw ~ "after",
                                year < ssw ~ "before",
                                TRUE ~ NA),
         ais = case_when(!is.na(zm) & is.na(ssw) ~ "zm",
                         !is.na(ssw) & is.na(zm) ~ "ssw",
                         !is.na(ssw) & !is.na(zm) ~ "both",
                         is.na(ssw) & is.na(zm) ~ "none")) 
age_data  %>% 
  group_by(ais) %>% 
  summarise(n_distinct(lake_id))
#we see here we have 
```

#now let's filter more closely for analysis 
-we want pre and post zebra mussel
  -filter for age 0-3 fish and sampling in zebra mussel lakes must be at least three years after        invasion
-we will assign pre/post to uninvaded systems based on the nearest zebra mussel lake spatially
```{r}
#add exposure to zebra mussels in the fish's life
age_data <- age_data %>% 
  filter(ais == "zm") %>% 
  mutate(birth_year = year - est_age,
         zm_lag = zm + 2,
         #we want fish that were captured before invasion AND fish born after invasion
         exposure = case_when(year <= zm_lag ~ "before",
                              birth_year >= zm_lag ~ "after",
                              TRUE ~ "mixed")) 

#how many fish do we have in each category 
age_data %>% 
  group_by(species_1) %>% 
  summarise(n_fish_before = sum(exposure == "before"),
            n_fish_after = sum(exposure == "after"),
            n_fish_mixed = sum(exposure == "mixed"))

age_data %>% 
  group_by(lake_id, species_1, year, zm_lag, exposure) %>% 
  summarise(n_fish = n(),
            n_age_classes = n_distinct(est_age))


#filter for fish that only have before and after (est ages could be based on mixed cohorts before and after)
less_restrictive <- age_data %>% 
  group_by(lake_id, species_1) %>% 
  summarise(
    has_before = any(exposure == "before"),
    has_after  = any(exposure == "after"),
    .groups = "drop"
  ) %>% 
  filter(has_before & has_after)

#now lets do a more restrictive filtering for only when an alk contains pre and post fish (no mixing)
#the only change here is that I filter for places where only age exists and then do the filtering 
actual_age_data <- age_data %>% 
  filter(ais == "zm" & !is.na(age)) %>% 
  mutate(birth_year = year - age,
         zm_lag = zm + 2,
         #we want fish that were captured before invasion AND fish born after invasion
         exposure = case_when(year <= zm_lag ~ "before",
                              birth_year >= zm_lag ~ "after",
                              TRUE ~ "mixed")) 

#two step fitler here that removes lake-speices-years that are mixed across the invasion year
more_restrictive <- actual_age_data %>% 
   group_by(lake_id, species_1, year) %>% 
  summarise(
    has_before = any(exposure == "before"),
    has_after  = any(exposure == "after"),
    has_mixed = any(exposure == "mixed"),
    .groups = "drop"
  ) %>% 
  filter(has_mixed == F) %>% 
  group_by(lake_id, species_1) %>% 
  summarise(
    has_before = any(has_before == T),
    has_after  = any(has_after == T),
    .groups = "drop"
  ) %>%
  filter(has_before & has_after) 
```
*I stopped right here on 3/24: I did two different filters. The first filters at the est age level so that we do not include fish that have an estimated age that spans the zebra mussel invasion, but the ALK could be constructed of a cohort of fish that lived in both conditions. My second filtering looks at the actual observed ages, making sure that the birth years do not cross the invasion line, such that the ALK is constructed in such a way that the fish only have lived in either condition. I still need to gather those fish. For the more restrictive grouping I will need to make sure that I don't still pick up the mixed groups if there was a mixed year and an after year in the lake-species*
