---
title: "growth_analysis_clean"
output: html_document
date: "2026-04-15"
editor_options: 
  chunk_output_type: console
---

# library
```{r}
library(tidyverse)
library(arrow)
library(ggplot2)
library(dplyr)
library(sf)
library(ggspatial)
library(ggthemes)
library(rnaturalearth)
library(rnaturalearthdata)
library(leaflet)
library(brms)
#speeds up model fitting
library(cmdstanr)
options(brms.backend = "cmdstanr")
```

# data
```{r}
#reading in the updated Minnesota data
mn_data <- open_dataset(sources = file.path("D:",
                                            "Shared drives",
                                            "Hansen Lab",
                                            "RESEARCH PROJECTS",
                                            "Fish Survey Data",
                                            "Parquet files",
                                            "mn_update")) %>% 
  filter(state == "Minnesota" & !is.na(est_age))

zm_ssw <- read_csv(file.path("D:",
                             "Shared drives",
                             "Hansen Lab",
                             "RESEARCH PROJECTS",
                             "MAISRC Mercury Pathways 53.2",
                             "Data",
                             "zm_ssw_6jan2026.csv")) %>% 
  #filter out na dows or those "none"
  filter(!is.na(dow) & !(str_detect(dow, "n"))) %>% 
  #generate original id for reference
  mutate(original_id = dow) %>% 
  #remove dashes
  #make sure short dows are 8 digits
  #fix a few manually
  #remove leading 0s to match the fish data
  mutate(dow = str_replace_all(dow, pattern = "-", ""),
         dow = if_else(nchar(dow) == 6, paste0(dow, "00"), dow),
         dow = case_when(dow == "10179" ~ "01017900",
                         str_detect(dow, "040063") ~ "04006300",
                         TRUE ~ dow),
         dow = case_when(str_starts(dow, "0") ~ str_sub(dow, 2),
                         TRUE ~ dow))  %>% 
  #rename dow to match fishereis column
  rename(lake_id = dow) %>% 
  #pivot to lake per row, with zm and starry columns 
  pivot_wider(id_cols = c(lake_id, lake_name),
              names_from = ais,
              values_from = year_infested,
              values_fn = min) %>% 
  rename(zm = `zebra mussel`,
         ssw = `starry stonewort`)

#crosswalk - get lat and lon from fish data
xwalk <- open_dataset(sources = file.path("D:",
                                            "Shared drives",
                                            "Hansen Lab",
                                            "RESEARCH PROJECTS",
                                            "Fish Survey Data",
                                            "Parquet files",
                                            "mn_update")) %>% 
  select(state, latitude_lake_centroid, lake_id, longitude_lake_centroid) %>% 
  filter(state == "Minnesota") %>% 
  filter(!is.na(latitude_lake_centroid) & !is.na(lake_id)) %>% 
  distinct(lake_id, latitude_lake_centroid, longitude_lake_centroid) %>% 
  collect()
```

# filter for analysis 
```{r}
age_data <- mn_data %>% 
  #filter MN data for fish with alked year level data
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  select(lake_id,
         nhdhr_id,
         date_total_effort_ident,
         date_sample,
         year,
         sampling_method,
         site_id,
         species_1,
         length_1,
         length_unit_1,
         age,
         aging_structure_1,
         sex,
         est_age,
         alk,
         alk_age_str,
         alk_length,
         alk_n,
         weight_1,
         weight_unit_1,
         flag,
         notes_1,
         zm) %>% 
  collect() %>% 
  #only take zebra mussel lakes
  filter(!is.na(zm)) %>% 
  #make timing columns
  mutate(zm_lag = zm+2,
         birth_year = year - est_age,
         exposure = case_when(
           year < zm_lag ~ "before",
           birth_year >= zm_lag ~ "after",
           TRUE ~ "drop"
         )) %>% 
  #remove mix fish
  filter(exposure != "drop")

valid_groups <- age_data %>% 
  group_by(lake_id, species_1) %>% 
  summarise(
    has_before = any(exposure == "before"),
    has_after = any(exposure == "after"),
    .groups = "drop"
  ) %>% 
  filter(has_before & has_after)

#gather fish that have before and after data
age_data_filtered <- age_data %>% 
  semi_join(valid_groups, by = c("lake_id", "species_1"))


# now lets filter for a minimum number of unique age classes within each group
age_check <- age_data_filtered %>%
  group_by(lake_id, species_1) %>%
  summarise(
    n_ages_before = n_distinct(est_age[exposure == "before" & !is.na(est_age)]),
    n_ages_after  = n_distinct(est_age[exposure == "after"  & !is.na(est_age)]),
    .groups = "drop"
  ) %>%
  filter(n_ages_before >= 5, n_ages_after >= 5)

age_data_filtered <- age_data_filtered %>% 
  semi_join(age_check, by = c("lake_id", "species_1"))

#quick check
check <- age_data_filtered %>% 
  group_by(lake_id, species_1) %>% 
  summarise(n_fish_before = sum(exposure == "before"),
            n_fish_after = sum(exposure == "after"),
            n_ages_before = n_distinct(est_age[exposure == "before"]),
            n_ages_after = n_distinct(est_age[exposure == "after"]))
rm(check, age_check, age_data, valid_groups)
#we see here that all lake-species-exposure have at least five different ages

#now create objects for each species 
species_list <- split(age_data_filtered, age_data_filtered$species_1)
```
This filtering is created on the estimated age, where we do not accept fish that were born before invasion but captured after (mixed). HOWEVER, if an ALK is created with mixed fish, that effect would then make it to the fish. We should be more careful to filter for where aged fish do not cross the line and then collect back all est aged fish that are from those alks

# model
```{r}
#lets walleye first
wae <- species_list[["walleye"]]

#make sure we only used common ages across lakes
#first we look at ages that are the same in both exposures in each lake
ages_within_lake <- wae %>%
  group_by(lake_id, est_age) %>%
  summarise(
    has_before = any(exposure == "before"),
    has_after  = any(exposure == "after"),
    .groups = "drop"
  ) %>%
  filter(has_before & has_after)

#now look at ages across all lakes
ages_global <- ages_within_lake %>%
  group_by(est_age) %>%
  summarise(
    n_lakes = n_distinct(lake_id),
    .groups = "drop"
  )
#looks reasonable to keep ages 1:7 (throws out a few lakes but maximizes est_ages and # of lakes)

#################stopped right here, I need to filter for cases where ages exsits before and after AND across all lakes, it was breaking for some reason right here... looks like some "drop" exposure was coming back########
#I think i found why - I had exposure spelt wrong in the fitler above - rerun and check
required_ages <- 2:5

valid_lakes <- wae %>%
  filter(est_age %in% required_ages) %>%
  group_by(lake_id, exposure) %>%
  summarise(
    n_ages = n_distinct(est_age),
    min_age = min(est_age),
    max_age = max(est_age),
    .groups = "drop"
  ) %>% 
  filter(n_ages == 4, min_age == 2, max_age == 5) %>% 
  group_by(lake_id) %>%
  filter(n_distinct(exposure) == 2) %>% 
  pull(lake_id)

wae_filtered <- wae %>%
  filter(lake_id %in% valid_lakes,
         est_age %in% required_ages)

wae_filtered %>% 
  group_by(lake_id, exposure) %>% 
  summarise(n_ages =  n_distinct(est_age),
            min = min(est_age),
            max = max(est_age)) %>% 
  print(n = nrow(.))

fit_wae <- brm(length_1 ~ est_age * exposure + (1|lake_id),
               data = wae,
               family = gaussian(),
               chains = 3,
               iter = 5000,
               warmup = 1000)
```

