---
title: "growth_analysis"
author: "Denver Link"
date: "2025-06-23"
output: html_document
editor_options: 
  chunk_output_type: console
---

# Packages
```{r}
library(tidyverse)
library(arrow)
library(ggplot2)
library(dplyr)
library(sf)
library(ggspatial)
library(ggthemes)
library(rnaturalearth)
library(rnaturalearthdata)
library(leaflet)
library(brms)
#speeds up model fitting
library(cmdstanr)
options(brms.backend = "cmdstanr")
```

# Data
-growth data from MN
-zm and ssw lakes from mn infested waters list
```{r}
#reading in the updated Minnesota data
mn_data <- open_dataset(sources = file.path("D:",
                                            "Shared drives",
                                            "Hansen Lab",
                                            "RESEARCH PROJECTS",
                                            "Fish Survey Data",
                                            "Parquet files",
                                            "mn_update")) %>% 
  filter(state == "Minnesota" & !is.na(est_age))

zm_ssw <- read_csv(file.path("D:",
                             "Shared drives",
                             "Hansen Lab",
                             "RESEARCH PROJECTS",
                             "MAISRC Mercury Pathways 53.2",
                             "Data",
                             "zm_ssw_6jan2026.csv")) %>% 
  #filter out na dows or those "none"
  filter(!is.na(dow) & !(str_detect(dow, "n"))) %>% 
  #generate original id for reference
  mutate(original_id = dow) %>% 
  #remove dashes
  #make sure short dows are 8 digits
  #fix a few manually
  #remove leading 0s to match the fish data
  mutate(dow = str_replace_all(dow, pattern = "-", ""),
         dow = if_else(nchar(dow) == 6, paste0(dow, "00"), dow),
         dow = case_when(dow == "10179" ~ "01017900",
                         str_detect(dow, "040063") ~ "04006300",
                         TRUE ~ dow),
         dow = case_when(str_starts(dow, "0") ~ str_sub(dow, 2),
                         TRUE ~ dow))  %>% 
  #rename dow to match fishereis column
  rename(lake_id = dow) %>% 
  #pivot to lake per row, with zm and starry columns 
  pivot_wider(id_cols = c(lake_id, lake_name),
              names_from = ais,
              values_from = year_infested,
              values_fn = min) %>% 
  rename(zm = `zebra mussel`,
         ssw = `starry stonewort`)

#crosswalk - get lat and lon from fish data
xwalk <- open_dataset(sources = file.path("D:",
                                            "Shared drives",
                                            "Hansen Lab",
                                            "RESEARCH PROJECTS",
                                            "Fish Survey Data",
                                            "Parquet files",
                                            "mn_update")) %>% 
  select(state, latitude_lake_centroid, lake_id, longitude_lake_centroid) %>% 
  filter(state == "Minnesota") %>% 
  filter(!is.na(latitude_lake_centroid) & !is.na(lake_id)) %>% 
  distinct(lake_id, latitude_lake_centroid, longitude_lake_centroid) %>% 
  collect() 
```

# Exploring growth data
```{r}
#how many lakes do we have for each species at the year level?
mn_data %>% 
  filter(alk == "year") %>% 
  group_by(species_1) %>% 
  summarise(n_lakes = n_distinct(lake_id)) %>% 
  collect()

#how many lakes with lake-year data do we have for each invasive?
mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  group_by(lake_id) %>% 
  mutate(ais = case_when(!is.na(zm) & is.na(ssw) ~ "zm",
                         !is.na(ssw) & is.na(zm) ~ "ssw",
                         !is.na(ssw) & !is.na(zm) ~ "both",
                         is.na(ssw) & is.na(zm) ~ "none")) %>% 
  ungroup() %>% 
  distinct(lake_id, ais) %>% 
  group_by(ais) %>% 
  count() %>% 
  collect()

#what is the general sense for species?
check <- mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  group_by(lake_id, species_1) %>% 
  mutate(ais = case_when(!is.na(zm) & is.na(ssw) ~ "zm",
                         !is.na(ssw) & is.na(zm) ~ "ssw",
                         !is.na(ssw) & !is.na(zm) ~ "both",
                         is.na(ssw) & is.na(zm) ~ "none"),
         impact_zm = case_when(year >= zm ~ "after",
                               year < zm ~ "before",
                               TRUE ~ "NA"),
         impact_ssw = case_when(year >= ssw ~ "after",
                                year < ssw ~ "before",
                                TRUE ~ "NA")) %>% 
  ungroup() %>% 
  distinct(lake_id, ais, species_1, impact_zm, impact_ssw) %>% 
  collect()
#we can see here if we have before and after for each invasive - NA for where it is uninvaded for that species
glimpse(check)

#general on fish we have growth data
invasion_summary <- check %>%
  group_by(lake_id, species_1) %>%
  summarise(
    zm_before = any(impact_zm == "before"),
    zm_after  = any(impact_zm == "after"),
    ssw_before = any(impact_ssw == "before"),
    ssw_after  = any(impact_ssw == "after"),
    .groups = "drop"
  ) %>%
  mutate(
    zm_status = case_when(
      zm_before & zm_after ~ "both",
      zm_before & !zm_after ~ "before only",
      !zm_before & zm_after ~ "after only",
      !zm_before & !zm_after ~ "neither"
    ),
    ssw_status = case_when(
      ssw_before & ssw_after ~ "both",
      ssw_before & !ssw_after ~ "before only",
      !ssw_before & ssw_after ~ "after only",
      !ssw_before & !ssw_after ~ "neither"
    )
  )

invasion_counts <- invasion_summary %>%
  group_by(species_1) %>% 
  count(zm_status, ssw_status) %>% 
  pivot_longer(cols = c(zm_status, ssw_status),names_to = "ais",values_to = "timing") %>% 
  group_by(species_1, ais, timing) %>% 
  summarise(n = sum(n))

invasion_counts %>% 
  filter(timing != "neither") %>% 
  ggplot() +
  geom_col(aes(timing, n)) +
  facet_grid(species_1~ais)

invasion_counts %>% 
  filter(timing != "neither") %>% 
  #filter(ais == "zm_status") %>% 
  ggplot() +
  geom_col(aes(timing, n)) +
  facet_grid(ais~species_1)

#how many ssw lakes do we have pre and post?
invasion_counts %>% 
  filter(ais == "ssw_status" & timing == "both")

ssw_lakes <- check %>% 
  filter(ais %in% c("ssw", "both")) %>% 
  select(-impact_zm, -ais) %>% 
  distinct(lake_id, species_1, impact_ssw) %>% 
  mutate(indicator = 1) %>% 
  pivot_wider(names_from = impact_ssw,
              values_from = indicator)
glimpse(ssw_lakes)

ssw_lakes_both <- ssw_lakes %>% 
  filter(before == "1" & after == "1") %>% 
  select(-after, -before) %>% 
  group_by(lake_id) %>% 
  summarise(species_1 = list(species_1), .groups = "drop") %>% 
  mutate(species_1 = sapply(species_1, toString))
  

###years of data?
years <- mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  group_by(lake_id, species_1, year) %>% 
  mutate(ais = case_when(!is.na(zm) & is.na(ssw) ~ "Zebra Mussel",
                         !is.na(ssw) & is.na(zm) ~ "Starry Stonewort",
                         !is.na(ssw) & !is.na(zm) ~ "ZM and SSW",
                         is.na(ssw) & is.na(zm) ~ "Uninvaded")) %>% 
  distinct(lake_id, ais, species_1, year) %>% 
  collect()

years %>% 
  group_by(species_1, lake_id, ais) %>% 
  count() %>% 
  ggplot() +
  geom_histogram(aes(n)) +
  labs(x = "# of years with halk at lake-year level",
       y = "# of lakes") +
  facet_grid(species_1~ais, scales = "free")

#how far after invasion for zm lakes?
before_after_lakes <- mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  filter(!is.na(zm) & is.na(ssw)) %>% 
  distinct(lake_id, species_1, year, zm) %>% 
  collect() %>% 
  mutate(status = case_when(zm <= year ~ "after",
                            TRUE ~ "before")) %>% 
  group_by(lake_id, species_1) %>% 
  summarise(has_before = any(status == "before"),
            has_after = any(status == "after")) %>% 
  filter(has_before == T & has_after == T) 

years <- mn_data %>% 
  right_join(before_after_lakes, by = c("lake_id", "species_1")) %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  mutate(status = case_when(zm <= year ~ "after",
                            TRUE ~ "before")) %>% 
  mutate(impact_years = year-zm) %>% 
    collect() %>% 
  group_by(lake_id, species_1) %>% 
  summarise(n_fish_before = sum(status == "before"),
            n_after = sum(status == "after"),
            n_after_less_5 = sum(status == "after" & impact_years <=5),
            n_after_5_plus = sum(status == "after" & impact_years >=5))

#how many lakes have before and after data, with after data at least 5 years removed from impact?
years %>% 
  filter(n_after_5_plus > 0) %>% 
  group_by(species_1) %>% 
  summarise(n_lakes = n_distinct(lake_id))
#we can see here we are really limited in data of lakes that were sampled 5 years after invasion and have data pre and post
```

# Map (filtering for zebra mussel lakes with before and after data)
*In the future this should be moved after filtering to represent the data that are being modeled*
   -this will involve retaining some objects created for the following chunk
```{r}
lake_species <- mn_data %>% 
  filter(alk == "year") %>% 
  left_join(zm_ssw, by = "lake_id") %>% 
  group_by(lake_id, species_1) %>% 
  mutate(ais = case_when(!is.na(zm) & is.na(ssw) ~ "zm",
                         !is.na(ssw) & is.na(zm) ~ "ssw",
                         !is.na(ssw) & !is.na(zm) ~ "both",
                         is.na(ssw) & is.na(zm) ~ "uninvaded"),
         impact_zm = case_when(year >= zm ~ "after",
                               year < zm ~ "before",
                               TRUE ~ "NA")) %>% 
  ungroup() %>% 
  distinct(lake_id, ais, species_1, impact_zm) %>% 
  collect() %>% 
  pivot_wider(id_cols = c(lake_id, species_1, ais), values_from = impact_zm,names_from = impact_zm) %>% 
  #zm_ssw has two a record for both zm and ssw for these lakes - leading an erroneous uninvaded row
  filter(!(lake_id == "18031100" & ais == "uninvaded") &
           !(lake_id == "4007900" & ais == "uninvaded")) %>% 
  #filter for cases where uninvaded lakes have data and invaded systems have pre and post
  mutate(timing = case_when(ais == "uninvaded" ~ "y",
                            ais == "ssw" ~ "n",
                            ais == "both" ~ "n",
                            ais == "zm" & !is.na(after) & !is.na(before) ~ "y",
                            TRUE ~ "n")) %>% 
  filter(timing == "y")

#we now attach spatial data, remove where they are na because they are mostly river or small
lake_species <- lake_species %>% 
  left_join(xwalk) %>% 
  filter(!is.na(latitude_lake_centroid)) %>% 
  select(lake_id, species_1, ais, latitude_lake_centroid, longitude_lake_centroid) %>% 
  #clean up species names 
  mutate(species_clean = case_when(species_1 == "black_crappie" ~ "Black Crappie",
                               species_1 == "bluegill" ~ "Bluegill",
                               species_1 == "cisco" ~ "Cisco",
                               species_1 == "largemouth_bass" ~ "Largemouth Bass",
                               species_1 == "northern_pike" ~ "Northern Pike",
                               species_1 == "smallmouth_bass" ~ "Smallmouth Bass",
                               species_1 == "walleye" ~ "Walleye",
                               species_1 == "yellow_perch" ~ "Yellow Perch"))

lake_sf <- lake_species %>%
  distinct(lake_id, species_clean, ais, latitude_lake_centroid, longitude_lake_centroid) %>%
  st_as_sf(coords = c("longitude_lake_centroid", "latitude_lake_centroid"), crs = 4326)

us_states <- ne_states(country = "United States of America", returnclass = "sf")
mn_border <- us_states %>% filter(name == "Minnesota")

# Plot
lake_uninvaded <- lake_sf %>% filter(ais == "uninvaded")
lake_zm <- lake_sf %>% filter(ais == "zm")

ggplot() +
  geom_sf(data = mn_border, fill = NA, color = "black") +
  geom_sf(data = lake_uninvaded, color = "lightblue", alpha = 0.5, size = 1) +
  geom_sf(data = lake_zm, color = "salmon", alpha = 0.75, size = 1) +
  facet_wrap(~ species_clean) +
  theme_minimal() +
  labs(color = "Invasion") +
  coord_sf(xlim = st_bbox(mn_border)[c("xmin", "xmax")],
           ylim = st_bbox(mn_border)[c("ymin", "ymax")]) +
  theme(axis.text.x = element_text(angle = 30, hjust = 1))
#ggsave("all_spec_map.jpg", height = 5, width = 5)

#subset
lake_species_sub <- lake_species %>% 
  left_join(xwalk) %>% 
  filter(!is.na(latitude_lake_centroid)) %>% 
  select(lake_id, species_clean, species_1, ais, latitude_lake_centroid, longitude_lake_centroid) %>% 
  filter(species_clean %in% c("Walleye", "Northern Pike", "Bluegill"))

lake_sf <- lake_species_sub %>%
  distinct(lake_id, species_clean, species_1, ais, latitude_lake_centroid, longitude_lake_centroid) %>%
  st_as_sf(coords = c("longitude_lake_centroid", "latitude_lake_centroid"), crs = 4326)

lake_uninvaded <- lake_sf %>% filter(ais == "uninvaded")
lake_zm <- lake_sf %>% filter(ais == "zm")

ggplot() +
  geom_sf(data = mn_border, fill = NA, color = "black") +
  geom_sf(data = lake_uninvaded, color = "lightblue", alpha = 0.5, size = 1) +
  geom_sf(data = lake_zm, color = "salmon", alpha = 0.75, size = 1) +
  facet_wrap(~ species_clean) +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        strip.text = element_text(size = 12),
        axis.text.x = element_text(angle = 30, hjust = 1)) +
  labs(color = "Invasion") +
  coord_sf(xlim = st_bbox(mn_border)[c("xmin", "xmax")],
           ylim = st_bbox(mn_border)[c("ymin", "ymax")])
#ggsave("filtered_spec_map.jpg", height = 3, width = 5)

rm(check,
   invasion_counts,
   invasion_summary,
   lake_sf,
   lake_species,
   lake_uninvaded,
   lake_zm,
   mn_border,
   ssw_lakes,
   ssw_lakes_both,
   us_states,
   years)
```

# ZM, WAE-NOP-BLG raw fish data for growth
```{r}
age_data.check <- mn_data %>% 
  #filter MN data for fish with alked year level data
  filter(alk == "year") %>% 
  #filtering join to the lake-species combos we want
  right_join(lake_species_sub, by = c("lake_id", "species_1"), suffix = c("_fish", "_invasion")) %>% 
  collect() %>% 
  #selecting columns important from the fish data that are relevant 
  select(lake_id,
         nhdhr_id,
         date_total_effort_ident,
         date_sample,
         year,
         sampling_method,
         site_id,
         species_1,
         length_1,
         length_unit_1,
         age,
         aging_structure_1,
         sex,
         est_age,
         alk,
         alk_age_str,
         alk_length,
         alk_n,
         weight_1,
         weight_unit_1,
         flag,
         notes_1) %>% 
  #bring in invasion data 
  left_join(zm_ssw, by = "lake_id") %>% 
  #create columns for invasion impact for both invaders
  mutate(impact_zm = case_when(year >= zm ~ "after",
                               year < zm ~ "before",
                               TRUE ~ NA),
         impact_zm_years = year - zm,
         impact_ssw = case_when(year >= ssw ~ "after",
                                year < ssw ~ "before",
                                TRUE ~ NA),
         ais = case_when(!is.na(zm) & is.na(ssw) ~ "zm",
                         !is.na(ssw) & is.na(zm) ~ "ssw",
                         !is.na(ssw) & !is.na(zm) ~ "both",
                         is.na(ssw) & is.na(zm) ~ "none")) 
```

#now let's filter more closely for analysis 
-there are several ways we could filter to pull out the zm effect
1. create an exposure variable based on the estimated age of the fish, filter for fish that were sampled before invasion (before grouping) or birth year is after invasion (after)... do not keep fish with birth year before invasion but sampled after (mixed) 
*Problem: some cohorts in the ALKs that generate this data could straddle invasion with their life*
2. filter for cases when the entire cohort of the ALK is after ZM invasion
   -more restrictive, less lake-species 
```{r}
#add exposure to zebra mussels in the fish's life
age_data <- age_data %>% 
  filter(ais == "zm") %>% 
  mutate(birth_year = year - est_age,
         zm_lag = zm + 2,
         #we want fish that were captured before lag invasion AND fish born after invasion lag
         exposure = case_when(year <= zm_lag ~ "before",
                              birth_year >= zm_lag ~ "after",
                              TRUE ~ "mixed")) 

#how many fish do we have in each category 
age_data %>% 
  group_by(species_1) %>% 
  summarise(n_fish_before = sum(exposure == "before"),
            n_fish_after = sum(exposure == "after"),
            n_fish_mixed = sum(exposure == "mixed"))

age_data %>% 
  group_by(lake_id, species_1, year, zm_lag, exposure) %>% 
  summarise(n_fish = n(),
            n_age_classes = n_distinct(est_age))


#filter for fish that only have before and after (est ages could be based on mixed cohorts before and after)
less_restrictive <- age_data %>% 
  group_by(lake_id, species_1) %>% 
  summarise(
    has_before = any(exposure == "before"),
    has_after  = any(exposure == "after"),
    .groups = "drop"
  ) %>% 
  filter(has_before & has_after)

#now join back to fish level and filter out "fish that cross the line"
before_after_fish <- age_data %>% 
  inner_join(less_restrictive, by = c("lake_id", "species_1")) %>% 
  filter(exposure %in% c("before", "after"))

less_restrictive %>% 
  group_by(species_1) %>% 
  summarise(n_lakes = n_distinct(lake_id))

less_restrictive %>% 
  group_by(lake_id) %>% 
  summarise(n_spec = n_distinct(species_1)) %>% 
  filter(n_spec > 1) %>% 
  arrange(desc(n_spec))

fish_counts <- before_after_fish %>% 
  group_by(lake_id, species_1) %>% 
  summarise(n_before = sum(exposure == "before"),
            n_after = sum(exposure == "after"))

##################################################################################

#now lets do a more restrictive filtering for only when an alk contains pre and post fish (no mixing)
#the only change here is that I filter for places where only age exists and then do the filtering 
actual_age_data <- age_data %>%
  filter(ais == "zm" & !is.na(age)) %>%
  mutate(birth_year = year - age,
         zm_lag = zm + 2,
         #we want fish that were captured before invasion AND fish born after invasion
         exposure = case_when(year <= zm_lag ~ "before",
                              birth_year >= zm_lag ~ "after",
                              TRUE ~ "mixed"))

#clean lake-years
clean_years <- actual_age_data %>% 
  group_by(lake_id, species_1, year) %>% 
  summarise(
    has_before = any(exposure == "before"),
    has_after = any(exposure == "after"),
    has_mixed = any(exposure == "mixed"),
    .groups = "drop"
  ) %>% 
  filter(!has_mixed) %>% 
  mutate(exposure = case_when(
    has_before ~ "before",
    has_after ~ "after"
  ))
 
#clean lake-species
valid_lake_species <- clean_years %>% 
  group_by(lake_id, species_1) %>% 
  summarise(
    has_before = any(exposure == "before"),
    has_after = any(exposure == "after"),
    .groups = "drop"
  ) %>% 
  filter(has_before & has_after)

final_years <- clean_years %>%
  semi_join(valid_lake_species, by = c("lake_id", "species_1"))

only_before_after <- actual_age_data %>%
  inner_join(final_years, 
             by = c("lake_id", "species_1", "year"))


#APRIL 14 - it would be of interest to explore why we end up with so few lakes (including some long invaded systems like Mille Lacs), I did check where Mille Lacs gets weeded out, at the lake_species object creation - it is invaded in 2006 and the earliest age fish are from 2006. In that object, those fish are all treated as "after" and there for it gets weeded out becuase it doesn't have before data

#also it looks like in this product that all fish have been aged, we can connect back to est aged fish as needed, we really just need to walk back our filtering steps to make sure we get everything
```

# gam model fit
```{r}
only_before_after <- only_before_after %>% 
  filter(species_1 == "bluegill")

fit <- brm(length_1 ~ age*exposure + (1|lake_id),
           data = only_before_after)
summary(fit)
plot(conditional_effects(fit))
```


# vb model fit
```{r}
########write out fish growth formula for modeling###############
vb_formula <- bf(
  #linf = asymptoic length, exp(lk) = growth coefficient k, t0 = curve shift (theoretical age at length 0)
  alk_length ~ Linf * (1 - exp(-exp(lk) * (est_age - t0))),
  
  #Asymptotic length - each lake gets its own baseline growth and invasion response
  Linf ~ 1 + after + (1 | lake_id)
  ,
  #growth coefficient - each lake gets its own baseline growth and invasion response
  lk   ~ 1 + after + (1 | lake_id)
  ,
  #keep t0 shared across groups
  t0   ~ 0,
  nl = TRUE
)


#######################old model structure#######################
# vb_formula <- bf(
#   #linf = asymptoic length, exp(lk) = growth coefficient k, t0 = curve shift (theoretical age at length 0)
#   length_1 ~ Linf * (1 - exp(-exp(lk) * (est_age - t0))),
#   
#   #Asymptotic length - each lake gets its own baseline growth and invasion response
#   Linf ~ 1 + after + species_1 + (1 | lake_id)
#   ,
#   #growth coefficient - each lake gets its own baseline growth and invasion response
#   lk   ~ 1 + after + species_1 + (1 | lake_id)
#   ,
#   #keep t0 shared across groups
#   t0   ~ 1,
#   nl = TRUE
# )
#this was an old model structure, I also included species specific slopes in one run - these did not converge well - biggest difference is the species term and that t0 is not fixed at 0
########################################################



##############put on some priors##########################
priors <- c(
  # Intercepts (population-level)
  prior(normal(400, 100), class = "b", nlpar = "Linf", coef = "Intercept"), #mean of 50 ranging from 0-100, informative but within reason
  prior(normal(log(0.3), 0.3), class = "b", nlpar = "lk", coef = "Intercept"), #mean of 0.2 roughly (0.03-1.5), covers very slow and fast growth but limits to anything really insane
  #prior(normal(-0.5, 0.3), class = "b", nlpar = "t0", coef = "Intercept"), # most mass between -2 and 2

  # Fixed effects (after)
  prior(normal(0, 5), class = "b", nlpar = "Linf", coef = "after"), # after can shift roughly +/- 20 units, wide enough for expected biological affects
  prior(normal(0, 0.5), class = "b", nlpar = "lk", coef = "after"), # allows big biological effects but prevents wild over fitting in growth rate
  
  # Species effect - assume difference are pretty moderate between species 
  # prior(normal(0, 15), class = "b", nlpar = "Linf", coef = "species_1"),
  # prior(normal(0, 0.5), class = "b", nlpar = "lk", coef = "species_1"),
  

  # Group-level SDs (random effects)
  prior(exponential(0.1), class = "sd", nlpar = "Linf", group = "lake_id"), #intercept
  prior(exponential(2), class = "sd", nlpar = "lk", group = "lake_id"), #intercept
  
  #prior(exponential(1), class = "sd", nlpar = "Linf", group = "lake_id", coef = "after"), #slopes
  #prior(exponential(1), class = "sd", nlpar = "lk", group = "lake_id", coef = "after"), #slopes
  
  #correlation between random intercepts and slopes
  #prior(lkj(2), class = "cor", group = "lake_id"),

  # Residual
  # heavy tails, centered at 0 - allows residual SD up to 20-30 without any resistance
  prior(student_t(3, 0, 5), class = "sigma")
)

#######filter data for model fitting#########################
glimpse(before_after_fish)
#need year to be a factor
before_after_fish <- before_after_fish %>% 
  mutate(year = as.factor(year)) %>% 
  mutate(after = case_when(exposure == "before" ~ 0,
                           exposure == "after" ~ 1)) %>% 
  #for species specific run
  filter(species_1 %in% c("northern_pike"))

before_after_fish %>% 
  group_by(exposure, after) %>% 
  count()

before_after_fish %>% 
  summarise(n_lakes = n_distinct(lake_id))

before_after_fish %>% 
  group_by(lake_id) %>% 
  summarise(n_fish_before = sum(after == 0),
            n_fish_after = sum(after == 1)) %>% 
  print(n = nrow(.))

#there are some lakes with a really small after grouping, lets weed those out here
before_after_fish <- before_after_fish %>% 
  group_by(lake_id) %>% 
  filter(sum(after == 1) >= 40) %>% 
  ungroup() 

before_after_fish %>% 
  group_by(lake_id) %>% 
  filter(sum(after == 1) >= 40) %>% 
  ungroup() %>% 
  ggplot() +
  geom_point(aes(age, length_1, color = as.factor(after))) +
  geom_smooth(aes(age, length_1, color = as.factor(after))) +
  facet_wrap(~lake_id)

#######fit model########################
fit_vb <- brm(
  vb_formula,
  data = before_after_fish,
  prior = priors,
  chains = 4,
  cores = 4,
  iter = 3000,
  warmup = 1500,
  control = list(adapt_delta = 0.95, max_treedepth = 15),
  backend = "cmdstanr", 
  threads = threading(2) 
)
saveRDS(fit_vb, file.path("G:",
                          "Shared drives",
                          "Hansen Lab",
                          "RESEARCH PROJECTS",
                          "MAISRC Mercury Pathways 53.2",
                          "Growth Analysis",
                          "growth_fit_pike.rds"))

fit_vb <- readRDS(file.path("D:",
                          "Shared drives",
                          "Hansen Lab",
                          "RESEARCH PROJECTS",
                          "MAISRC Mercury Pathways 53.2",
                          "Growth Analysis",
                          "growth_fit_pike.rds"))

########explore convergence##############
plot(fit_vb)
summary(fit_vb)
```

Things we could try in the future:
-we could try to fit a lake specific slope with the intercept and slope uncorrelated (after || lake_id)


# generate curves before and after 
```{r}
# age grid starting at 0 increasing by 0.1 until the max value from the fitted data
age_grid <- seq(0, max(before_after_fish$est_age, na.rm = TRUE), by = 0.1)

#expand that for both before and after values
newdata <- expand.grid(
  est_age = age_grid,
  after = c(0,1)
)

# get posterior predicted draws of the expected value (not including obs noise)
# use posterior_epred for expected value of response; re_formula = NA for population-level only
epreds <- posterior_epred(fit_vb, newdata = newdata, re_formula = NA)

# epreds: draws x rows(newdata)
# summarize to mean and 95% CI
pred_summary <- as.data.frame(t(apply(epreds, 2, function(x) {
  c(mu = mean(x), lwr = quantile(x, 0.025), upr = quantile(x, 0.975))
})))
pred_summary <- bind_cols(newdata, pred_summary)

# plot
ggplot(pred_summary, aes(x = est_age, y = mu, color = factor(after), fill = factor(after))) +
  geom_ribbon(aes(ymin = `lwr.2.5%`, ymax = `upr.97.5%`), alpha = 0.2, color = NA) +
  geom_line(size = 1) +
  labs(color = "after", y = "Length (mm)", x = "Age") +
  scale_fill_manual("Invasion", values = c("lightblue", "salmon"), labels = c("Before", "After")) +
  scale_color_manual("Invasion", values = c("lightblue", "salmon"), labels = c("Before", "After")) +
  #geom_vline(xintercept = 3, linetype = "dashed") +
  #geom_hline(yintercept = 381, linetype = "dashed") +
  theme_minimal() +
  theme(legend.position = c(0.8, 0.1))
#ggsave("wae_wb_zm_growth.jpg", height = 5, width = 5)
```

# generate lengths at age 3 for before and after zm invasion
```{r}
nd0 <- data.frame(est_age = 3, after = 0)
nd1 <- data.frame(est_age = 3, after = 1)

pred0 <- posterior_epred(fit_vb, newdata = nd0, re_formula = NA) # draws x 1
pred1 <- posterior_epred(fit_vb, newdata = nd1, re_formula = NA)

diff_draws <- as.vector(pred1) - as.vector(pred0)
quantile(diff_draws, c(0.025, 0.5, 0.975))

#mean before
mean(as.vector(pred0))
#mean after
mean(as.vector(pred1))
```

#data filtering/structure viz
```{r}
# Data for bars (sampling years and cohort spans)
df <- data.frame(
  group = c("2011", "2016", "2020"),
  y = c(1, 2, 3),
  start = c(2006, 2011, 2015),  # 6 years back (ages 0–5)
  end   = c(2011, 2016, 2020)
)

# Mixed portion (part to filter out)
mixed <- data.frame(
  y = 2,
  start = 2011,
  end = 2013
)

# Plot
ggplot() +
  # Background shading
  annotate("rect", xmin = 2005, xmax = 2013, ymin = 0.5, ymax = 3.5,
           fill = "#9ecae1", alpha = 0.6) +
  annotate("rect", xmin = 2013, xmax = 2020, ymin = 0.5, ymax = 3.5,
           fill = "#fcbba1", alpha = 0.6) +

  # Bars (full cohorts)
  geom_rect(data = df,
            aes(xmin = start, xmax = end, ymin = y - 0.2, ymax = y + 0.2),
            fill = "#de6b48", color = "black") +

  # Before-only bar recolored
  geom_rect(data = subset(df, group == "2011"),
            aes(xmin = start, xmax = end, ymin = y - 0.2, ymax = y + 0.2),
            fill = "#2b6c8a", color = "black") +

  # Mixed filtered portion
  geom_rect(data = mixed,
            aes(xmin = start, xmax = end, ymin = y - 0.2, ymax = y + 0.2),
            fill = "#ffff33", color = "black") +

  # Invasion line
  geom_segment(
  aes(x = 2013, xend = 2013, y = 0.5, yend = 3.5),
  linewidth = 1 )+

  # Labels
  annotate("text", x = 2013, y = 3.7, label = "ZM Invasion\n2013", vjust = 0) +
  annotate("text", x = 2009, y = 3.7, label = "Before") +
  annotate("text", x = 2017, y = 3.7, label = "After") +
  annotate("text", x = 2012, y = 2, label = "Filtered Out", size = 3) +

  # Axis formatting
  scale_x_continuous(breaks = seq(2005, 2020, 1)) +
  scale_y_continuous(NULL, breaks = NULL) +

  # Theme
  theme_minimal(base_size = 14) +
  theme(
    panel.grid.major.y = element_blank(),
    panel.grid.minor = element_blank(),
    panel.grid.major.x = 
  ) +

  labs(
    x = "Year",
    caption = "*Each surveyed lake-year includes ages 0–5\n*Bars span birth years of observed cohorts"
  )
ggsave("data_structure.jpg", width = 7, height = 5)
```


